\plaintitle

### 2017-01-08 drafts/labor/complexity.tex
two attempts at pw perspective in complexity

{\topic{Only tasks that could be measured and priced could be completed via piecework
--- a fact which fundamentally limited the potential complexity of piecework.}
% When \citeauthor{Brown01041990} investigated
% what limited the adoption of piecework in industries that otherwise gravitated toward it
% (e.g., railway engineers),
% the homogeneity of tasks arose as a major contributing factor~\cite{Brown01041990}.
% \citeauthor{10.2307/23702539} concurs via a case study of the Santa Fe Railway,
% which used ``efficiency experts'' to develop a ``standard time''
% to determine pay for each task at the company informed by
% ``thousands of individual operations''% removed ``at the Topeka shop'' because even I had to reach back into the back of my mind to be like ``Oh, yeah, Kansas''.
% ~\cite{10.2307/23702539}.
Earlier we discussed \citeauthor{10.2307/23702539}'s and later \citeauthor{Brown01041990}'s analysis of railway workers;
one might conclude from \citeauthor{10.2307/23702539}'s observations in particular that
complex, creative work
--- which is inherently heterogeneous and difficult to routinize ---
would be unsuitable for piecework~\cite{10.2307/23702539}.
\citeauthor{Brown01041990}'s description of efficiency experts would corroborate this;
``efficiency experts'' can effectively gauge how long known tasks should take, but
would find themselves overwhelmed if they attempted to assess creative work like design,
which can take an arbitrary number of iterations before proceeding to a subsequent step
\ali{should i cite some design research here?}.

\topic{Piecework was limited to tasks that could be clearly evaluated.}
For example, the roles required to facilitate piecework in the early \nth{20} century included ``piecework clerks, inspectors, and `experts'''~\cite{10.2307/23702539}.
\citeauthor{hart2016rise} argue that evaluation is the ultimate complexity limit:
at some point, evaluating multidimensional work for quality
(rather than for quantity) becomes infeasible.
In their words,
``if the quality of the output is more difficult to measure than the quantity [\ldots]
then a piecework system is likely to encourage
an over--emphasis on quantity produced and an under--emphasis on quality''~\cite{hart2016rise}.
Complex work, which is often subjective to evaluate, falls victim to this criteria.

\topic{This focus on measurement and tracking had consequences.}
\citeauthor{10.2307/23702539} suggests that the first sparks of scientific management
could be found in piecework:
the approach of paying workers for each piece of output necessitated
the rigorous tracking, measurement, and training of workers
for which scientific management became famous~\cite{10.2307/23702539}.
If true, 
the concurrent upswing of
scientific management and Fordism
through the first two--thirds of the \nth{20} century
alongside piecework was not only understandable, but predictable~\cite{hart2013rise}.

\topic{Piecework researchers also argue that,
in addition to constraints on the kind of \textit{work} that's amenable to piecework,
only certain kinds of \textit{organizations} were amenable to piecework.}
Researchers detail three organizational criteria.
\begin{Numberlist}
\item \citeauthor{Brown01041990} argues that
piecework ``is less likely in jobs with a variety of duties than in
jobs with a narrow set of routinized duties''~\cite{Brown01041990}.
\citeauthor{SJOE:SJOE371} points out the phenomenon here as a market effect:
``in an environment with multi--tasking,
pay schemes based on tightly specified performance may
induce workers to neglect tasks that are less easy to measure''~\cite{SJOE:SJOE371}.
\item complexity was limited by access to capital to create the necessary infrastructure.
As \citeauthor{10.2307/23702539} reports,
only the largest and most wealthy railroads had the resources
necessary
% ``pay the overhead involved in installing work reorganization''
~\cite{10.2307/23702539}.
\item organizations required capable managers in charge of the pieceworkers.
The West Virginia mines, for example, hired foremen 
to be the intermediary between upper management and the workers~\cite{10.2307/2118435}.
These foremen were responsible for allocating resources and
understanding when and how to modify work as necessary~\cite{wray1949marginal}.
\end{Numberlist}
So, in sum,
organizations historically could only take advantage of piecework if they had
homogeneous work to be done,
access to capital to purchase the necessary equipment, and
the ability to hire people who could serve as intermediaries between pieceworkers and management.

\topic{The research seems to suggest that it was difficult to apply piecework to more skilled work,
particularly because maximizing the advantages of piecework seemed to reward
smaller, more constrained, more narrowly--trained tasks, and only in organizations that could pay for the equipment and people to enable it.}
For most of the \nth{19} century,
piecework was applied almost exclusively to farm and textile work.
Work was simple and widely understood
--- farm workers didn't need to be trained on how to plow fields, or birth foals;
seamstresses knew how to sew together denim~\cite{10.2307/2338394,riisOtherSideLives}.}

>>>>

\ali{What do we know?

Airy's computers:
\begin{itemize}
  \item we can do some pretty complex stuff if we break tasks down into quickly verifiable parts
  \item there's no need to train people to be experts for this work; non--experts can do the job just as well.
\end{itemize}

% Domestic and farm work
% \begin{itemize}
%   \item ???
% \end{itemize}

% match--girls
% \begin{itemize}
%   \item ???
% \end{itemize}

Industrial workers
\begin{itemize}
  \item revolutionary: we can train people on a narrow slice of the work and have them just do that work
\end{itemize}
}

\topic{Piecework's body of research most squarely addresses complexity in two of the cases we looked at earlier:
Airy's human computers and among industrial workers, especially during World War II.}
The majority of piecework's perspective, on this topic, will come from these areas.

\topic{Airy's work opened the door to greater task complexity by decomposition.}
Airy's computers had relatively limited education in mathematics, but
they were nevertheless collectively able to accomplish complex goals by
approaching distinct, subdivided, quickly verifiable tasks in chunks~\cite{grier2013computers}.
\ali{do i need to say more?}

\topic{It seemed, too, that only tasks that could be measured and priced could be completed via piecework
--- a conclusion which would fundamentally limit the potential complexity of piecework.}
Earlier we discussed \citeauthor{10.2307/23702539}'s and later \citeauthor{Brown01041990}'s analysis of railway workers;
one might conclude from \citeauthor{10.2307/23702539}'s observations in particular that
complex, creative work
--- which is inherently heterogeneous and difficult to routinize ---
would be unsuitable for piecework~\cite{10.2307/23702539}.
\citeauthor{Brown01041990}'s description of efficiency experts would corroborate this;
``efficiency experts'' can effectively gauge how long known tasks should take, but
would find themselves overwhelmed if they attempted to assess creative work like design,
which can take an arbitrary number of iterations before proceeding to a subsequent step.

\topic{The next innovation in piecework's drive toward complexity came on
the assembly line, specifically during the war effort.}
At a time when manufacturing firms needed more able--bodied workers, but
all of the conventionally trained metalworkers had joined the armed forces,
factories turned to women~\cite{honey1985creating,davies2014origins}.
This need to train workers rapidly to do very narrowly defined tasks resulted in
a \textit{recomposition} of work, wherein
workers were only prepared for the tasks they would do;
riveters would only be trained on how to do rivet and
to pass the output to the next person on the assembly line,
who could evaluate and send the work back or proceed with the next step.

\ali{OR\dots}

\topic{Only tasks that could be measured and priced could be completed via piecework
--- a fact which fundamentally limited the potential complexity of piecework.}
Earlier we discussed \citeauthor{10.2307/23702539}'s and later \citeauthor{Brown01041990}'s analysis of railway workers;
one might conclude from \citeauthor{10.2307/23702539}'s observations in particular that
complex, creative work
--- which is inherently heterogeneous and difficult to routinize ---
would be unsuitable for piecework~\cite{10.2307/23702539}.
\citeauthor{Brown01041990}'s description of efficiency experts would corroborate this;
``efficiency experts'' can effectively gauge how long known tasks should take, but
would find themselves overwhelmed if they attempted to assess creative work like design,
which can take an arbitrary number of iterations before proceeding to a subsequent step.

\topic{Piecework was limited to tasks that could be quickly and accurately evaluated.}
For example, the roles required to facilitate piecework
in the early \nth{20} century included ``piecework clerks, inspectors, and `experts'''~\cite{10.2307/23702539}.
\citeauthor{hart2016rise} argue that evaluation limited piecework's complexity:
at some point, evaluating multidimensional work for quality
(rather than for quantity) becomes infeasible.
In his words,
``if the quality of the output is more difficult to measure than the quantity [\ldots]
then a piecework system is likely to encourage
an over--emphasis on quantity~\dots~and an under--emphasis on quality''~\cite{hart2016rise}.

% \topic{This focus on measurement and tracking had consequences.}
% \citeauthor{10.2307/23702539} suggests that the first sparks of scientific management
% could be found in piecework:
% the approach of paying workers for each piece of output necessitated
% the rigorous tracking, measurement, and training of workers
% for which scientific management became famous~\cite{10.2307/23702539}.
% If true, 
% the concurrent upswing of
% scientific management and Fordism
% through the first two--thirds of the \nth{20} century
% alongside piecework was not only understandable, but predictable~\cite{hart2013rise}.

\topic{Piecework researchers were also finding that certain types of \textit{organizations},
as well as certain types of \textit{work},
were more amenable to piecework regimes.}
\citeauthor{Brown01041990} wrote that piecework was
``less likely in jobs with a variety of duties than in
jobs with a narrow set of routinized duties''~\cite{Brown01041990};
\citeauthor{SJOE:SJOE371} hypothesized this as an effect of market forces
--- suggesting that
``pay schemes based on~\dots~performance may induce workers to neglect tasks that are less easy to measure''~\cite{SJOE:SJOE371}.
Airy seemed to recognize this,
decomposing complex mathematical problems into easily verified components, % ~\cite{grier2013computers},
but applying this model to tasks like railway engineering may have been prohibitively difficult.

\topic{But it was more than the inherent qualities of work and the organization of the firm that mattered;
management practices themselves had to be appropriately geared to working with pieceworkers.}
Competent foremen were integral both for allocating resources and
for liaising between workers and upper management~\cite{wray1949marginal,10.2307/2118435}.
% The West Virginia mines, as \citeauthor{10.2307/2118435} writes,
% ~\cite{10.2307/2118435}.

% \topic{Piecework researchers also argued that,
% in addition to constraints on the kind of \textit{work} that's amenable to piecework,
% only certain kinds of \textit{organizations} were amenable to piecework.}
% Researchers detail three organizational criteria.
% First, \citeauthor{Brown01041990} found that
% piecework was ``less likely in jobs with a variety of duties than in
% jobs with a narrow set of routinized duties''~\cite{Brown01041990}.
% \citeauthor{SJOE:SJOE371} pointed out this phenomenon as a market effect:
% ``in an environment with multi--tasking,
% pay schemes based on tightly specified performance may
% induce workers to neglect tasks that are less easy to measure''~\cite{SJOE:SJOE371}.
% Second, complexity was limited by access to capital to create the necessary infrastructure.
% As \citeauthor{10.2307/23702539} reported,
% only the largest and most wealthy railroads had the resources
% necessary to change existing systems to piecework regimes~\cite{10.2307/23702539}.
% Third, organizations required capable managers in charge of the pieceworkers.
% The West Virginia mines, for example, hired foremen 
% to be the intermediary between upper management and the workers~\cite{10.2307/2118435}.
% These foremen were responsible for allocating resources and
% understanding when and how to modify work as necessary~\cite{wray1949marginal}.
% So, in sum,
% organizations historically could only take advantage of piecework if they had
% homogeneous work to be done,
% access to capital to purchase the necessary equipment, and
% the ability to hire people who could serve as intermediaries between pieceworkers and management.}

% \topic{The research seems to suggest that it was difficult to apply piecework to more skilled work,
% particularly because maximizing the advantages of piecework seemed to reward
% smaller, more constrained, more narrowly--trained tasks, and only in organizations that could pay for the equipment and people to enable it.}
% For most of the \nth{19} century,
% piecework was applied almost exclusively to farm and textile work.
% Work was simple and widely understood
% --- farm workers didn't need to be trained on how to plow fields, or birth foals;
% seamstresses knew how to sew together denim~\cite{10.2307/2338394,riisOtherSideLives}.
